RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-05 doc/bioinformatics2013/RVD2.tex
Paper draft update---modified derivation of parameter initialization using MOM

\subsection{Initialization}

The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure. Essentially, we set the parameter value (population moment) equal to the sample moment. In some cases, the method-of-moments estimate is outside the parameter space. Then, we XXX.

The estimated error probability is calculated from the observed sample
\begin{equation}
 \theta_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}

The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
\begin{equation}
 \mu_0=E(\mu_j|\mu_0,M_0)
\end{equation}
\begin{equation}
 M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
\end{equation}

@@

\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}

>>>>

\subsection{Initialization}
The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure. Essentially, we set the parameter value (population moment) equal to the sample moment. In some cases, the method-of-moments estimate is outside the parameter space. Then, we XXX.

The estimated error probability is calculated from the observed sample
\begin{equation}
 \tilde{\theta}_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}
The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 \tilde{\mu}_j =\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}
\begin{equation}
 \tilde{M}_j = \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
\end{equation}
Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
\begin{equation}
 \mu_0=E(\mu_j|\mu_0,M_0)
\end{equation}
\begin{equation}
 \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
\end{equation}

\begin{equation}
 M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
\end{equation}
\begin{equation}
 \tilde{M}_0 = \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
\end{equation}

@@

\section{Appendix}
\subsection{Parameter initialization derivation}
The model parameters and latent variables are initialized using method of moments by equating population moment to sample moment.

The estimated error probability is calculated from the observed sample
\subsubsection{$\theta_{ji}$}\
\begin{equation}
 \tilde{\theta}_{ji}=\frac{r_{ji}}{n_{ji}}
\end{equation}

\subsubsection{$\mu_j$ and $M_j$}\
\begin{align}
 \theta_{ji}|\mu_j,M_j\thicksim \text{Beta}(\mu_j,M_j)
\end{align}

The probability density function of beta distribution parameterized with the mean $\mu_j$ and sample-size paramenter $M_j$ is
\begin{equation}
\begin{split}
  f(\theta_{ji};\mu_j,M_j) = \frac{1}{B(M_j\mu_j,M_j(1-\mu_j))}\theta_{ji}^{M_j\mu_j-1}(1-\theta_{ji})^{M_j(1-\mu_j)-1}.
\end{split}
\end{equation}
The first polulation moment (mean) is
\begin{equation}
 \mu_j=\text{E}[\theta_{ji}],
\end{equation}
while the first sample moment
\begin{equation}
\begin{split}
 \text{E}[\tilde{\theta}_{ji}] &= \frac{1}{N-1}\sum_{i=1}^N\tilde{\theta}_{ji}\\ &=\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{split}
\end{equation}
Equat the population mean to sample mean we have
\begin{equation}
 \tilde{\mu}_j =\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\theta_{ji}) = \frac{\mu_j(1-\mu_j)}{M_j+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\theta}_{ji}) &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\text{E}[\tilde{\theta}_{ji}])^2\\ &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2
 \end{split}
\end{equation}
By definition of Method of moments,
\begin{equation}
  \frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\tilde{M}_j+1} = \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2,
\end{equation}
so,
\begin{equation}
\begin{split}
 \tilde{M}_j &= (N-1)\cdot\frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2}-1\\ &= (N-1)\cdot\frac{\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}\\ &= \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
\end{split}
\end{equation}

\subsubsection{$\mu_0$ and $M_0$}
\begin{align}
 \mu_j|\mu_0,M_0\thicksim \text{Beta}(\mu_0,M_0)
\end{align}
Likewise,
\begin{equation}
 \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
\end{equation}

The expected value ($\mu_0$) is
\begin{equation}
 \mu_0=\text{E}[\mu_j],
\end{equation}
while the first sample moment
\begin{equation}
\begin{split}
\text{E}[\tilde{\mu}_j] &= \frac{1}{J-1}\sum_{i=1}^J\tilde{\mu}_j\\ &=  \frac{1}{J-1}\sum_{i=1}^J\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\\ &=\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{split}
\end{equation}

Using Methond of Moment
\begin{equation}
  \tilde{\mu}_0 = \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\mu_j) = \frac{\mu_0(1-\mu_0)}{M_0+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\mu}_j) &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\text{E}[\tilde{\mu}_j])^2\\ &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2
 \end{split}
\end{equation}
By definition of Method of moments,
\begin{equation}
  \frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\tilde{M}_0+1} = \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2,
\end{equation}
so,
\begin{equation}
\begin{split}
 \tilde{M}_0 &= (J-1)\cdot\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2}-1\\ &= (J-1)\cdot\frac{\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1\\ &= \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
\end{split}
\end{equation}
\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}
